# Data Manipulation with dplyr

In this chapter, we will learn a very popular package ``dplyr`` to deal with data manipulation. We will mainly go through its main functions [@biomed1;@rpro1].

## select()

Let's first import the package and the ``mtcars`` dataset.

```{r message = FALSE}
install.packages(dplyr)
library(dplyr)
data(mtcars)
```

If we want to select some columns from the dataset. We could use the ``select()`` function in ``dplyr``. It is similar with the ``subset()`` function, but here, you do not need to use ``select = c('mpg', 'disp')``. You use the names of the columns directly in the function.

```{r}
df <- select(mtcars, # name of the data frame
       mpg, disp) # column names you want to select
head(df, 3)
```

You could also use the index of the columns.

```{r}
df <- select(mtcars, # name of the data frame
       c(1, 3)) # index of the columns you want to select
head(df, 3)
```

The codes above is kind of a *traditional* way to do the work. We start with a function and put parameters in the function. However, this is not the typical way to use ``dplyr``.

The codes below is a more ``dplyr`` way people use ``dplyr``. We start with the name of the data frame. Then, we put a speical sign ``%>%`` called ``pipe`` after it. We continue from a new line and write the function we want to use. Besides that, we could add more functions with the pipe operator. For example, only show first three observations with ``head()`` functions. 

```{r}
mtcars %>% # name of the data frame
  select(mpg, disp) %>% # select the columns by their names
  head(3)
```

We will keep using this fasion in the following lecutre.

Besides choosing some columns you want, you could also exclude the column you do not want by putting a negative sign ``-`` before the variable.

```{r}
mtcars %>%
  select(-mpg, -disp) %>%
  head(3)
```

You could use ``:`` to select a range of variables.

```{r}
mtcars %>%
  select(mpg:hp) %>% # select from mpg to hp in the data frame
  head(3)
```

## filter()

In ``dplyr``, we could use ``filter()`` function to select the rows satisfying some conditions.

```{r}
mtcars %>%
  filter(mpg > 30)
```

Add more conditions by using ``,`` to separate them.

```{r}
mtcars %>%
  filter(mpg > 30, qsec < 19)
```

## arrange()

We could arrange the order of some columns by ``arrange()`` functions.

```{r}
mtcars %>%
  arrange(mpg) %>% # arrange mpg in increasing order
  head(10)
```

Or maybe we want ``mpg`` to be in a descending order. Just put a ``desc()`` outside the variable.

```{r}
mtcars %>%
  arrange(desc(mpg)) %>%
  head(10)
```

We could put them together by using pipe operators to connect them.

```{r}
mtcars %>%
  select(gear, mpg) %>%
  arrange(desc(gear), mpg) %>%
  head(10)
```

## mutate()

We use ``mutate()`` to do some calculations within the variables and create a new column to store them.

```{r}
mtcars %>%
  select(mpg) %>%
  mutate(kmpg = mpg * 1.609) %>%
  head(10)
```

```{r}
mtcars %>%
  select(mpg, wt) %>%
  mutate(kmpg = mpg * 1.609, lbwt = wt * 1000) %>%
  head(10)
```

## group_by() and summarise()

We use ``group_by()`` to do aggregation (group the observations based the values of one or one more columns) work and ``summarise()`` to calculate some statistics related to each group.

```{r}
mtcars %>%
  group_by(gear) %>%
  summarise(mean_mpg = mean(mpg),
            max_mpg = max(mpg),
            total = n())
```

```{r}
mtcars %>%
  group_by(gear, am) %>%
  summarise(mean_mpg = mean(mpg),
            max_mpg = max(mpg),
            total = n())
```

## join()

We could use ``join()`` to do the same work of ``merge()``.

```{r}
ID <- c(1:4) # create variable ID
Name <- c('A', 'B', 'C', 'D') # create variable Name
Score1 <- c(69.5, 77.5, 99, 90) # create variable Score1
df1 <- data.frame(ID, Name, Score1) # combine the varibles into one data frame called df1
df1

Name <- c('A', 'D', 'C') # create variable Name
Score2 <- c(98, 46, 55) # create variable Score2
df2 <- data.frame(Name, Score2) # combine the varibles into one data frame called df2
df2
```

```{r warning = F}
df1 %>%
  left_join(df2, by = 'Name')
```

```{r warning = F}
df1 %>%
  right_join(df2, by = 'Name')
```

Could you tell the difference between ``left_join()`` and ``right_join()``?

Besides ``left_join()`` and ``right_join()``, we have ``inner_join()`` (keep only matched observations of two data frames) and ``full_join()`` (keep all observations of two data frames).




