# Data Visualization with ggplot2

In this chapter, we will learn to use ``ggplot2`` to do data visualization tasks. Before the lecture, please install the package first.

```{r eval=FALSE}
install.packages("ggplot2")
```


## Grammer of Graphics (gg)
We have grammar for languages. We also have grammars for graphcis. That's where ``gg`` of ``ggplot2`` comes from. For ``ggplot2``, it has seven grammatical elements listed in the table below [@datacamp1].

| Element     | Description                                       |
|-------------|---------------------------------------------------|
| Data        | The dataset being plotted.                        |
| Aesthetics  | The scales onto which we map our data.            |
| Geometries  | The visual elements used for our data.            |
| Facets      | Plotting small multiples.                         |
| Statistics  | Representations of our data to aid understanding. |
| Coordinates | The space on which the data will be plotted.      |
| Themes      | All non-data ink.                                  |

Let's take the codes below as a simple example to show how those different elements work in ggplots. From this example, we might have a general idea what is each element and how does it work. We will talk about each element in details later.

```{r warning = F, message = FALSE}
library(ggplot2)
data(mtcars)
ggplot(mtcars, # Data
       aes(x = mpg, y = wt)) + # Aesthetics
  geom_point() + # Geometries
  facet_grid(. ~ gear) + # Facets
  stat_smooth(method = "lm", se = FALSE, col = "blue") + # Statistics
  scale_x_continuous('Miles/(US) gallon',
                     limits = c(0, 40)) + 
  scale_y_continuous('Weight (1000 lbs)',
                     limits = c(0, 7)) + # Coordinates
  theme_bw() # Themes
```


## Data, Aesthetics, and Geometries

Generally, if you want to draw figures with ``ggplot2``, you need at least three elements, which are data, aesthetics, and geometries. Data is the dataset we want to visualize. Aesthetic specifies the variables and related attributes. Geometry indicates the plot type and related attributes. Take the example above again. We want to visualize the varialbes of ``mpg`` and ``wp`` (aesthetic) of the dataset ``mtcars`` (data) with a scatter plot (geometry).

Similar to ``dplyr``, ``ggplot2`` also has its own fashion of coding. We start with the ``ggplot()`` function. **Please pay attention that there is no ``2`` in the name of the function**. In the function, we first indicate the name of the dataset or data frame. Then we use ``aes()`` to indicate the scales we want to map our data. Here, we map ``mpg`` to x axis, and ``wt`` to y axis. Then we use a plus sign ``+`` to connect it to other functions. We are going to draw a scatter plot, so we use ``geom_point()``.

```{r}
ggplot(mtcars, # Data
       aes(x = mpg, y = wt)) + # Aesthetics
  geom_point() # Geometries
```

We could see the result and it is following our codes. Based on this plot, we have a overall idea of the relationship between ``wt`` and ``mpg``, which is ``wt`` has a negative relationship with ``mpg``. This result makes sense to us. If a car is lighter and it could drive farther with the same amount of gasoline. (Please remember that this relationship is just a type of correlation, not causality.)

We could add attributes in the aesthetic element. For example, we could use color to indicate the value of ``hp`` by adding ``col = hp``.

```{r}
ggplot(mtcars,
       aes(x = mpg, y = wt, col = hp)) +
  geom_point()
```

Now we have more information in the result. While the weight is heavier, the hourse power is stronger.

Here, ``hp`` is a continous variable, so ``ggplot2`` uses the darkness of the color to indicate the value. However, if we use a categorical varibable (e.g. binomial variable), ``ggplot2`` will use different colors to show different types.

```{r}
ggplot(mtcars,
       aes(x = mpg, y = wt, col = factor(am))) +
  geom_point()
```

Here, ``am`` stands for the types of transmission system (0 = automatic, 1 = manual). We use ``factor()`` to transfer this variable to a categorical one. Then, the ``ggplot2`` uses one color for automatic transmission and another color for mannual transmission.

Besides ``color``, there are several other parameters to show different aspects of the plots.

| Parameter | Description                             | Continuous variable | Categorical variable |
|-----------|-----------------------------------------|---------------------|----------------------|
| x         | X axis position                         | ✓                   |                      |
| y         | y axis position                         | ✓                   |                      |
| size      | Diameter of points, thickness of lines  | ✓                   |                      |
| alpha     | Transparency                            | ✓                   | ✓                    |
| color     | Color of dots, outlines of other shapes | ✓                   | ✓                    |
| fill      | Fill colour                             | ✓                   | ✓                    |
| labels    | Text on a plot or axes                  |                     | ✓                    |
| shape     | Shape of point                          |                     | ✓                    |
| linetype  | Line dash pattern                        |                     | ✓                    |

As for geometries, there are many different types ot them you can use for different plots. For examples, ``geom_point()`` for scatter plot, ``geom_bar`` for bar plot, ``geom_boxplot`` for boxplot, etc. Most functions of geometries are self-explained, so you could tell what their usages easily. We all talk about those commonly used geometries such as sactter plot, bar plot, line plot, etc. in the following parts.

### Scatter plot

We use ``geom_point()`` to plot scatter plot in R with ``ggplot2``. In the example below, we map ``mpg`` to x axis, and ``wt`` to y axis. We indicate the transparency by set ``alpha = hp``. The transparency of each point is decided by its value of ``hp``.

```{r}
ggplot(mtcars,
       aes(x = mpg, y = wt, alpha = hp)) +
  geom_point()
```

### Bar plot

In the example below, we draw a bar plot to show the number of cars with different tansmission systems. In ``aes()``, we only indicate the variable ``am``. R then will count the number for each transimission type. We use ``geom_bar()`` to plot it.

```{r}
ggplot(mtcars,
       aes(am)) +
  geom_bar()
```

Go back to our previous example, which is different from the example above.

```{r}
year <- c(1998:2003) # create variable year
sales <- c(500, 600, 650, 700, 400, 550) # create variable sales
df <- data.frame(year, sales) # combine the variables into one data frame called df

ggplot(df, aes(x = year, y = sales)) +
  geom_bar(stat = 'identity') # you need to specify stat = 'identity' to plot the actual value for each year, not count
```


### Line plot

To show the usage of line plot in ``ggplot2``, we use a new dataset ``economics`` coming from the ``ggplot2`` package. This dataset was produced from US economic time series data. Use ``help()`` to check more information of this dataset.

In this example, we use ``geom_line()``.

```{r}
data(economics)
ggplot(economics,
       aes(x = date, y = unemploy)) +
  geom_line(col = 'Blue') # indicate the color of the line by setting col = 'Blue'
```

In this plot, it shows clearly the temperol trend of the number of unemployed in the US. There two increase recently starting from 2001 and 2008.

### Boxplot

In the example below, we draw a boxplot for each transmission type. To do this, we map ``am`` to the x axis, and ``mpg`` to the y axis. We use ``factor()`` or ``as.factor()`` to transfer ``am`` to a factor (or categorical variable). 

```{r}
data(mtcars)
ggplot(mtcars, aes(x = as.factor(am), y = mpg)) + 
  geom_boxplot()
```

### Pie chart

In ``ggplot2``, it is not as intuitive as the base function ``pie()`` to draw a pie chart. That's use the same example to see the difference.

```{r}
year <- c(1998:2003) # create variable year
sales <- c(500, 600, 650, 700, 400, 550) # create variable sales
df <- data.frame(year, sales) # combine the variables into one data frame called df

ggplot(df, aes(x = '', y = sales, fill = factor(year))) +
  geom_bar(width = 1, stat = 'identity') +
  coord_polar('y')
```

What does ``ggplot2`` do here to draw a bar plot is to create a bar plot first.

```{r}
ggplot(df, aes(x = '', y = sales, fill = factor(year))) +
  geom_bar(width = 1, stat = 'identity')
```

And then transfer the coordinate system to the polar one.

```{r}
ggplot(df, aes(x = '', y = sales, fill = factor(year))) +
  geom_bar(width = 1, stat = 'identity') +
  coord_polar('y')
```

## Facets

If you want to split up your data by one or more variables and plot each subset in one figure, ``facet`` is the element you want to use.

In the following example, we draw a scatter plot for each number of forward gears. In each plot, we map ``mpg`` to the x axis and ``wt`` to y axis. The three plots are alligned in a row.

```{r message=FALSE}
data(mtcars)
ggplot(mtcars, # Data
       aes(x = mpg, y = wt)) + # Aesthetics
  geom_point() + # Geometries
  facet_grid(. ~ gear) # Facets, for each number of foward gears
```

If we want to allign the plots in a column.

```{r}
data(mtcars)
ggplot(mtcars, # Data
       aes(x = mpg, y = wt)) + # Aesthetics
  geom_point() + # Geometries
  facet_grid(gear ~ .) # Facets, pay attention to the position of gear and the dot sign
```

We could put more variables to split the plots. In the following example, we put one more variable ``vs`` in the ``facet_grid()``.

```{r}
ggplot(mtcars, # Data
       aes(x = mpg, y = wt)) + # Aesthetics
  geom_point() + # Geometries
  facet_grid(vs ~ gear) # Facets
```

We could use the ``margins = T`` to add more plots showing the aggregation of the plots in each column or row.

```{r}
ggplot(mtcars, # Data
       aes(x = mpg, y = wt)) + # Aesthetics
  geom_point() + # Geometries
  facet_grid(vs ~ gear, margins = T) # Facets
```

We could use ``labeller=label_both`` to add more information in the label.

```{r}
ggplot(mtcars, # Data
       aes(x = mpg, y = wt)) + # Aesthetics
  geom_point() + # Geometries
  facet_grid(vs ~ gear, labeller=label_both) # Facets
```


## Statistics

You could fit a line to see the general trend of the scatter plot by adding ``stat_smooth()``.

```{r}
p <- ggplot(mtcars,
       aes(x = mpg, y = wt)) +
  geom_point()
p +  stat_smooth(method = "lm") # linear line
```

```{r}
p + stat_smooth(method = 'loess') # non-linear line
```

You could use also use draw a histogram for the dataset.

```{r}
p <- ggplot(mtcars, aes(mpg))
p + geom_histogram(binwidth = 5) # the width of each category is 5
```

```{r}
p +  geom_histogram(bins = 3) # the number of bins is 3
```

You could fit a denisty line for the histogram.

```{r}
p + geom_histogram(aes( y = ..density..), binwidth = 3, col = 'Black', fill = 'White') +
  geom_density(alpha=.2, fill="Gray") 
```

Another way to show more statistics about your data is boxplot, which we have been introduced before.

```{r}
p <- ggplot(mtcars, aes(x = factor(gear), y = mpg))
p + geom_boxplot()
```


## Coordinates

While there are many coordinate systems supported by ``ggplot2``, the most commonly used is Cartesian coordinate system, which is the combination of x axis and y axis orthogonally.

### Zooming in and out

In the following example, we zoom in our plot to a spcific area.

```{r}
data(mtcars)
p <- ggplot(mtcars, aes(x = mpg, y = wt)) +
      geom_point()
p + coord_cartesian(xlim = c(10, 30), ylim = c(2,4))
```

### Ratio

We could change the ratio of the length of a y unit relative to the length of a x unit ($\frac{\text{y unit}}{\text{x unit}}$).

```{r}
ggplot(mtcars, # Data
       aes(x = mpg, y = wt)) + # Aesthetics
  geom_point() +
  coord_fixed(ratio = 1)
```

```{r}
ggplot(mtcars, # Data
       aes(x = mpg, y = wt)) + # Aesthetics
  geom_point() +
  coord_fixed(ratio = 5)
```

### Swaping the axes

```{r}
ggplot(mtcars, # Data
       aes(x = mpg, y = wt)) + # Aesthetics
  geom_point() +
  coord_flip()
```

```{r}
ggplot(mtcars, aes(x = as.factor(am), y = mpg)) + 
  geom_boxplot() +
  coord_flip()
```


```{r}
ggplot(mtcars, # Data
       aes(gear)) + # Aesthetics
  geom_bar() +
  scale_y_reverse()
```

### Polar coordinate system

We touched the polar coordinate system a little bit when drawing a pie chart.

```{r}
ggplot(mtcars, aes(x = '', fill = factor(gear))) +
  geom_bar(width = 1) +
  coord_polar(theta = 'y')
```


```{r}
ggplot(mtcars, aes(factor(gear))) +
  geom_bar(width = 1, col = 'Black', fill = 'Grey') +
  coord_polar()
```

## Themes

``ggplot2`` is powerful in its flexibility of themes.

### Add labels

```{r}
p <- ggplot(mtcars, aes(x = mpg, y = wt)) +
  geom_point()
p + labs(title = 'Plot 1',
         x = 'Miles/(US) gallon',
         y = 'Weight (1000 lbs)')
```

```{r}
p + ggtitle('Plot1') +
  xlab('Miles/(US) gallon') +
  ylab('Weight (1000 lbs)')
```

### Change ticks

```{r}
p <- ggplot(mtcars, aes(x = factor(am), y = mpg))
p + geom_boxplot() +
    scale_x_discrete(name = "Transmission", 
                    labels = c('Automatic', 'Manual'))
```

### theme() function

``theme()`` function could help to change the styles of every components of the plots. Here are a few examples about it [@ggplot1].

```{r}
p <- ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point() +
  labs(title = 'Fuel economy declines as weight increases')
p
```

We could use ``plot.title`` to change the style of a title in the plot. In the following exmaple, we change the font size of the title by setting ``element_text()`` and making ``size`` to be larger as the defual one.

```{r}
p + theme(plot.title = element_text(size = rel(2)))
```

We could also use absolute value to indicate the size directly.

```{r}
p + theme(plot.title = element_text(size = 15))
```

We could change the background of the plot by setting the value of ``plot.background`` in the ``theme()`` function. For example, if we want to change the color, we could set ``element_rect()`` and ``fill = 'red'``.

```{r}
p + theme(plot.background = element_rect(fill = 'red'))
```

More specifically, if we want to change the style of the panel, which is the inner part restricted by x and y axes, we could set the value of ``panel.background``.

```{r}
p + theme(panel.background = element_rect(fill = 'green', colour = 'red'))
```

We could set the line type of the panel's border.

```{r}
p + theme(panel.border = element_rect(linetype = 'dashed', fill = NA))
```

```{r}
p + theme(panel.grid.major = element_line(colour = "black"),
          panel.grid.minor = element_line(colour = 'blue'))
```

```{r}
p + theme(panel.grid.major.y = element_blank(),
          panel.grid.minor.y = element_blank())
```

We could put the gridline on the top of our data.

```{r}
p + theme(
  panel.background = element_rect(fill = NA),
  panel.grid.major = element_line(colour = 'Blue'),
  panel.ontop = TRUE
)
```

```{r}
p + theme(
  axis.line = element_line(size = 3, colour = 'red')
)
```

```{r}
p + theme(
  axis.text = element_text(colour = 'Green', size = 15)
)
```

```{r}
p + theme(
  axis.ticks = element_line(size = 1.5)
)
```

```{r}
p + theme(
  axis.title.y = element_text(size = rel(1.5), angle = 30)
)
```

Now, let's see what we could do for the legend.

```{r}
p <- ggplot(mtcars, aes(wt, mpg)) +
  geom_point(aes(colour = factor(cyl), shape = factor(vs))) +
  labs(
    x = 'Weight (1000 lbs)',
    y = 'Fuel economy (mpg)',
    colour = 'Cylinders',
    shape = 'Transmission system'
  )
p
```

```{r}
p + theme(
  legend.position = 'none'
)
```

```{r}
p + theme(
  legend.position = 'top'
)
```

```{r}
p + theme(
  legend.justification = 'top'
)
```

```{r}
p + theme(
  legend.position = c(.95, .95),
  legend.justification = c('right', 'top'),
  legend.box.just = 'right'
)
```

```{r}
p + theme(
  legend.box.background = element_rect(),
  legend.box.margin = margin(6, 6, 6, 6)
)
```

```{r}
p + theme(
  legend.key = element_rect(fill = 'white', colour = 'black')
)
```

```{r}
p + theme(
  legend.text = element_text(size = 8, colour = 'red', face = 'bold')
)
```
