# Data Visualization with Base Functions

We would focus on the base plotting functions in R first and then switch to the ggplot2 part.

## Scater plot
```{r}
year <- c(1998:2003) # create variable year
sales <- c(500, 600, 650, 700, 400, 550) # create variable sales
df <- data.frame(year, sales) # combine the variables into one data frame called df
plot(df)
```

Or, you could use the variable names directly and indicate the dataset as the codes below. You will get the same result.
```{r}
plot(sales ~ year, data = df)
```


## Line plot
You could transfer the scatter plot above to a line plot by just adding a ``type`` variable to indicate that you are plotting a line.
```{r}
plot(df,
     type = 'l') # type indicates the line type
```

You could also choose another type by changing the value of ``type``, as the one below.
```{r}
plot(df,
     type = 'b') # type indicates the line type
```

## Bar plot
```{r}
barplot(df$sales,
        names.arg = df$year) # names.org indicates the vector of names to be plotted under each bar
```

## Add more elements in the plots
For a reader-friendly plots, you have to add more information such as titile, labels, and legend. For the plot above, we could use the codes below to make it more informative.
```{r}
barplot(df$sales,
        names.arg = df$year,
        main = 'Bar plot of the sales for each year from 1998 to 2003', # add title for the plot
        xlab = 'Year', # add label tag for the x-axis
        ylab = 'Sales (million dollors)',  # add label for the y-axis
        legend = 'Sales') # add lengend name
```