# Statistics in R

## simple statistics
### Mean and median
We could use the base functions to do some simple statistical analysis directly.
```{r}
a <- c(10, 20, 30, 50, 33, 29, 10, 30, 100, 0, 4.5)
mean(a) # mean
median(a) # median
```

### Minimum and maximum value
```{r}
min(a) # minimum value
max(a) # maximum value
```

### Quantiles
```{r}
x <- quantile(a)
x # list of quantiles
x[2] # select the value by its index
```

## Simple Linear regression
Our example is the dataset ``mtcars``, and we want to explore the relationship between ``mpg`` (Miles/(US) gallon) and ``wt`` (Weight (1000 lbs)).
```{r}
data(mtcars)
library(ggplot2)
ggplot(mtcars, aes(mpg, wt)) +
  geom_point()
```

It seems there is a linear relationship between these two variables. We then add a linear line to fit them.
```{r}
ggplot(mtcars, aes(mpg, wt)) +
  geom_point() +
  geom_smooth(method = "lm", se = F)
```

We then use the ``lm()`` function to fit this linear relationship and use ``summary()`` function to see the result.
```{r}
lm_fit <- lm(wt ~ mpg,      # formula
             mtcars)        # dataset
summary(lm_fit)             # check result
summary(lm_fit)$r.squared   # value of R2
coefficients(lm_fit)        # only check the coefficient
```

## Multiple Regression
Most of the time, we need to examine the relationship between one dependent variable with several independent variables. Then we have carry out multiple regression. Here, ``multiple`` indicates more than one independent variables.
```{r}
mlm_fit <- lm(mpg~disp+hp+wt,
          mtcars)
summary(mlm_fit)
summary(mlm_fit)$r.squared
```

## Logistic regression
The above two examples both use continuous variables as their dependent variables. How about using a binomial variable a dependent variable? Then we need to do logistic regression. There are many functions to do this. Here, we introduced the ``glm()`` function.
```{r}
logit_reg <- glm(formula = am ~ cyl + hp + wt,
                 mtcars,
                 family = binomial)
summary(logit_reg)
```


**Reference**

1. [R - Mean, Median and Mode](https://www.tutorialspoint.com/r/r_mean_median_mode.htm)

